A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-06 tex/sec5-synthesis.tex
[ tex ] Fix undefined references
[ tex ] Minor revisions up to S4
[ tex ] Some progress in S5

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}

\subsection{Mode Correctness}
\begin{definition}[Mode-Correctness]
\end{definition}

\begin{theorem}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  Then, for every raw term $\isTerm{t}$ in synthesising mode and two derivations of 
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
\end{theorem}

\subsection{Trichotomy of Raw Terms by Type Synthesis}

@@

\begin{corollary}[Trichotomy of Raw Terms]
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  Then, for every context $\Gamma$ and raw term $\isTerm{t}$, exactly one of the following statements holds:
  \begin{enumerate}
    \item $\isTerm{t}$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation for some type $A$.
    \item $\isTerm{t}$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has no derivation for any type $A$.
    \item $\isTerm{t}$ is not synthesising.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis}.

@@

\subsection{Decidability of Bidirectional Type Synthesis and Checking}

\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  For every context $\Gamma$ and raw term $\isTerm{t}$ in mode $d$, 
  \begin{enumerate}
    \item if $\isTerm{t}$ is synthesising, then it is decidable whether there is a type~$A$ and a derivation of
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
      \]
    \item if $\isTerm{t}$ is checking, then every type $A$ decides whether there is a derivation of
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A.
      \]
  \end{enumerate}
\end{theorem}
\begin{lemma}\label{lem:args-induction}
  Assume \Cref{thm:bidirectional-type-synthesis}.
  Then, for every set $\Xi$ of type variables, list 
  \[
    \mathit{as} = \left([\Delta_i] A_{i}^{\dir{d_i}}\right)_{i = 1}^{n}
  \]
  list of raw terms $t_i$'s in mode $d_i$, context $\Gamma$, and \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
  \begin{enumerate}
    \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
      \[
        \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
      \]

    \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation. 
  \end{enumerate}
\end{lemma}

@@

  \end{description}
\end{proof}

>>>>

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
This section is devoted to defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
\Cref{subsec:mode-correctness} formalises the concept of mode-correctness and establishes the uniqueness of synthesised types.
This indicates that for a given raw term and context, any two synthesised types must be equal.
This characteristic is also used in the proof of the logical decidability of bidirectional type synthesis.
Later, in \Cref{subsec:trichotomy}, we introduce the decidability of bidirectional synthesis but postpone its proof until \Cref{subsec:bidirectional-synthesis-checking}.
At that point, we will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to show the trichotomy on raw terms as a simple corollary.
Note that the result is applicable to an arbitrary mode-correct bidirectional type system, rather than being restricted to the simply typed $\lambda$-calculus as we did in \Cref{sec:key-ideas}.

\subsection{Mode Correctness}\label{sec:mode-correctness}
As \citet[Section~3.1]{Dunfield2021} outlined, mode-correctness for a bidirectional type system means that for every bidirectional typing rule:
\begin{enumerate*}
\item each `input' type variable in a premise must be synthesised from `earlier' premises or, if the mode of the conclusion is checking, provided by the conclusion;
\item each `output' type variable in the typing judgement of the conclusion should have been synthesised from premises already, if the mode of the conclusion is synthesis.
\end{enumerate*}
An `input' type variable refers to types in an extension context and the type of a premise in the checking mode.
It is important to note that the order of premises in a bidirectional typing rule matters, as it also sets the order of bidirectional type synthesis.

Consider the rule $\ChkRule{Abs}$ (\Cref{fig:STLC-bidirectional-typing-derivations}) as an example.
This rule is mode-correct, as the type variables $A$ and $B$ in its only premise are already provided by its conclusion $A \bto B$.
Likewise, the rule $\SynRule{App}$ for an application term $\isTerm{t\;u}$ is mode-correct because:
\begin{enumerate*}
\item the type $A \bto B$ of the first argument $t$ is synthesised, thereby ensuring type variables $A$ and $B$ must be known if successfully synthesised;
\item the type of the second argument $u$ is checked against $A$, which was synthesised earlier;
\item as a result, the type of an application $t\;u$ can be synthesised.
\end{enumerate*}

As this is a condition for \emph{typing rules} and not for raw terms or typing derivations, this property \emph{cannot} be formalised without the concept of bidirectional binding signature presented in \Cref{sec:defs}.
This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without the general definition of bidirectional type systems.
Now that we have the notion of signature available, we can articulate mode-correctness more precisely.

First, let us introduce the finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.
}
$\fv(A)$ and $\fv(\Gamma)$ of (free) variables in a type and in a context, respectively, which are defined in the usual way:
\begin{enumerate*}
\item For a type $A : \Type_{\Sigma}(\Xi)$ define the subset $\fv(A)$ of $\Xi$ inductively by $\fv(x) = {x}$ and $\fv(\tyOp_{i}(A_1, \ldots, A_n)) = \bigcup_{i = 1}^{n} \fv(A_i)$;
\item by reusing this notation, for a context $\Gamma : \Type_{\Sigma}(\Xi)$ define the set $\fv(\Gamma)$ as $\fv(\emptyset) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
\end{enumerate*}
As we have sketched informally, the mode-correctness condition for a synthesis rule is different from a checking rule
\begin{definition}[Mode-Correctness] \label{def:mode-correctness}
  An operation $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}}}{A_0^{\dir{d}}}$ is \emph{mode-correct} whenever
  \begin{enumerate}
    \item if $d$ is $\syn$ (synthesis);
    \item if $d$ is $\chk$ (checking);
  \end{enumerate}

\end{definition}

\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  Then, for every raw term $\isTerm{t}$ in synthesising mode and two derivations of 
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
\end{lemma}

\subsection{Trichotomy on Raw Terms by Type Synthesis} \label{subsec:trichotomy}

@@

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  Then, for every context $\Gamma$ and raw term $\isTerm{t}$, exactly one of the following statements holds:
  \begin{enumerate}
    \item $\isTerm{t}$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation for some type $A$.
    \item $\isTerm{t}$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has no derivation for any type $A$.
    \item $\isTerm{t}$ is not synthesising.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine  \Cref{thm:soundness,thm:completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.

@@

\subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}

\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  For every context $\Gamma$ and raw term $\isTerm{t}$ in mode $d$, 
  \begin{enumerate}
    \item if $\isTerm{t}$ is synthesising, then it is decidable whether there is a type~$A$ and a derivation of
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
      \]
    \item if $\isTerm{t}$ is checking, then every type $A$ decides whether there is a derivation of
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A.
      \]
  \end{enumerate}
\end{theorem}

@@

  \begin{claim}\label{lem:args-induction}
    For every set $\Xi$ of type variables, list 
    \[
      \mathit{as} = \left([\Delta_i] A_{i}^{\dir{d_i}}\right)_{i = 1}^{n}
    \]
    list of raw terms $t_i$'s in mode $d_i$, context $\Gamma$, and \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
    \begin{enumerate}
      \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
        \[
          \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
        \]

      \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation. 
    \end{enumerate}
  \end{claim}
  \end{description}
\end{proof}
